Score-based tests of measurement invariance: Use in practice

### 2014-03-24 frontiers/frontiers.Rnw
set up presentation.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@322 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor changes to frontiers manuscript.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@332 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

To verbally describe Equation~\eqref{eq:score}, each
individual has $k$ scores describing the extent to which the fitted
model describes that particular individual.  We can roughly interpret
scores near zero as providing a ``good'' description of an individual,
with scores far from zero providing a ``bad'' description of an
individual.  This is only a rough interpretation as, even when measurement
invariance holds, some individuals' scores will be further from zero
than others.  However, under measurement noninvariance, the scores
will differ for different subgroups of individuals (say, scores in
subgroup A tend to be negative and scores in subgroup B tend to be
positive).  Each of the $k$ scores represents one model parameter,
which, as further described below, allows us to test subsets of model
parameters for invariance.  While scores can be obtained
under the multivariate normal likelihood (discrepancy) function and
alternatives such as generalized least squares, most SEM software
fails to supply the scores to the user.

@@

In this section, we demonstrate how the above tests can be carried out in \proglang{R}, 
using the package \pkg{lavaan} \cite{lavaan11} for model estimation and \pkg{strucchange} 
\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
the applicability of adult gratitude scales to youth, available in the~\proglang{R} 
package \pkg{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
gratitude scales from $n=1401$ youth aged 10--19 years. 
The original authors 
were specifically interested in whether the scales were measurement invariant 
across age.  Because the sample size at each age was unbalanced, the authors 
created age groups of approximately equal sample size.  In the examples below, 
we test for measurement invariance across these age groups.
For illustrative purposes, we conduct multiple tests and
compare them to the traditional significance level of 0.05.  In
practice, however, one should generally adjust the significance level
for the number of tests carried out.

@@

full <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
  data = yg, group = "agegroup", meanstructure = TRUE)
@
%
Finally, we can get the results of a likelihood ratio test via the \code{anova()}
function, which implies that the GQ-6 violates measurement invariance.
<<LRT>>=
anova(full, restr)
@
To obtain the asymptotically equivalent $\mathit{LM}_\mathit{uo}$ (Equation~\eqref{eq:lmuo}), we
can use the \code{sctest()} function from \pkg{strucchange}:
<<LM_UO>>=
sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
  functional = "LMuo")
@
This command specifies that we assess the parameter 1--4 of
model \code{restr} after ordering the observations according to
\code{agegroup}.  Additionally, 
the observed information matrix is used as the variance-covariance matrix. Note
that the model parameters 1--4 are the factor loadings supplied by
\pkg{lavaan}, which can be seen by inspecting \code{coef(restr)}. This
also leads to somewhat 
smaller test statistics that are very close to being significant at the 5\% level.

@@

In this paper, we focused on testing for measurement invariance in factor
analysis models that assume multivariate normality and that are estimated via
maximum likelihood (ML).  The family of tests described here generally apply to estimation methods
that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are
potentially applicable to alternative SEM discrepancy functions such
as generalized 
least squares \cite<e.g.,>{broarm95}.  Score calculation for these alternative
discrepancy functions has not been implemented (to our knowledge), though the
calculation could be implemented.  Test statistic calculation and
inference would then proceed in exactly the same manner as the calculation and
inference illustrated in this paper.

In addition to alternative fit functions, the tests can be extended to other
models estimated via ML.  Of primary interest to the topic of measurement
invariance, the tests can be extended to item response models to examine
differential item functioning.  In particular, \citeA{strkop13} studied
application of these tests to the Rasch model, using them as the basis of a
recursive partitioning procedure that segments subgroups of individuals who
exhibit DIF.  Further study and extension of these tests for IRT are warranted.

>>>>

To verbally describe Equation~\eqref{eq:score}, each
individual has $k$ scores describing the extent to which the fitted
model describes that particular individual.  
These scores are similar to residuals \cite<and, in fact, the tests
can be applied directly to residuals in other contexts; see>{zeihor07}:
we can roughly interpret
scores near zero as providing a ``good'' description of an individual,
with scores far from zero providing a ``bad'' description of an
individual.  This is only a rough interpretation as, even when measurement
invariance holds, some individuals' scores will be further from zero
than others.  However, under measurement noninvariance, the scores
will differ for different subgroups of individuals (say, scores in
subgroup A tend to be negative and scores in subgroup B tend to be
positive).  Each of the $k$ scores represents one model parameter,
which, as further described below, allows us to test subsets of model
parameters for invariance.  While scores can be obtained
under the multivariate normal likelihood (discrepancy) function and
alternatives such as generalized least squares, most SEM software
fails to supply the scores to the user.

@@

In this section, we demonstrate how the above tests can be carried out in \proglang{R}, 
using the package \pkg{lavaan} \cite{lavaan11} for model estimation and \pkg{strucchange} 
\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
the applicability of adult gratitude scales to youth, available in the~\proglang{R} 
package \pkg{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
gratitude scales from $n=1401$ youth aged 10--19 years. 
The original authors 
were specifically interested in whether the scales were measurement invariant 
across age.  Because the sample size at each age was unbalanced, the authors 
created age groups of approximately equal sample size.  In the examples below, 
we test for measurement invariance across these age groups.
For illustrative purposes, we conduct multiple tests and
compare them to the traditional significance level of 0.05.  In
practice, however, one should generally adjust the significance level
for the number of tests carried out.  Additionally, because
measurement invariance researchers often have large sample sizes,
cross-validation methods can be useful to help verify the test
results.

@@

full <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
  data = yg, group = "agegroup", meanstructure = TRUE)
@

Finally, we can get the results of a likelihood ratio test via the \code{anova()}
function, which implies that the GQ-6 violates measurement invariance.
<<LRT>>=
anova(full, restr)
@
To obtain the asymptotically equivalent $\mathit{LM}_\mathit{uo}$ (Equation~\eqref{eq:lmuo}), we
can use the \code{sctest()} function from \pkg{strucchange}:
<<LM_UO>>=
sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
  functional = "LMuo")
@
This command specifies that we assess the parameter 1--4 of
model \code{restr} after ordering the observations according to
\code{agegroup}.  Additionally, 
the observed information matrix is used as the variance-covariance matrix. Note
that the model parameters 1--4 are the factor loadings supplied by
\pkg{lavaan}, which can be seen by inspecting \code{coef(restr)}. This
also leads to somewhat 
smaller test statistics that are very close to being significant at the 5\% level.

@@

In this paper, we focused on testing for measurement invariance in factor
analysis models that assume multivariate normality and that are estimated via
maximum likelihood (ML).  The family of tests described here generally apply to estimation methods
that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are
potentially applicable to alternative SEM discrepancy functions such
as generalized 
least squares \cite<e.g.,>{broarm95}.  Score calculation for these alternative
discrepancy functions has not been implemented (to our knowledge), though the
calculation could be implemented.  Test statistic calculation and
inference would then proceed in exactly the same manner as the calculation and
inference illustrated in this paper.  Study of the proposed tests'
application to larger SEMs is warranted.

In addition to alternative fit functions, the tests can be extended to other
models estimated via ML.  Of primary relevance to the topic of measurement
invariance, the tests can be extended to item response models to examine
differential item functioning.  In particular, \citeA{strkop13} studied
application of these tests to the Rasch model, using them as the basis of a
recursive partitioning procedure that segments subgroups of individuals who
exhibit DIF.  Further study and extension of these tests for IRT are warranted.
